Bibat: Batteries-include Bayesian Analysis Template

### 2024-02-26 paper/automl-template.tex
Update paper

@@

@@

@@

@@

@@

>>>>

    \answerNA{I can't think of any particular negative societal impacts of a Bayesian workflow template}

@@

    \answerYes{See \url{https://bibat.readthedocs.io/en/latest/_static/report.html} for instructions to reproduce the main example.}

@@

    \answerYes{See section "Generating Stan inputs" here \url{https://bibat.readthedocs.io/en/latest/_static/report.html\#preparing-the-data}}

@@

    \answerNo{This is unnecessary as the results are easy to reproduce and the results are large files that would be awkward to store online}

@@

    \answerNo{To reproduce the figures from raw data run "make analysis" from the folder \texttt{example\_projects/baseball}, as described in the instructions here \url{https://bibat.readthedocs.io/en/latest/\_static/report.html}}

@@

    \answerYes{\url{https://github.com/teddygroves/bibat/}}

### 2024-02-26 paper/main.qmd
Update paper

The term "Bayesian workflow" captures the idea that Bayesian statistical
analysis comprises not just inference, but also specific approaches to related
activities like data preparation, model design, diagnosis, debugging and
criticism. This idea can be found in @boxBayesianInferenceStatistical1992
and has recently received increasing scholarly recognition of
[@gelmanBayesianWorkflow2020;@grinsztajnBayesianWorkflowDisease2021;
@gabryVisualizationBayesianWorkflow2019]. Software tools now exist
that address most individual aspects of a Bayesian workflow: see
@strumbeljPresentFutureSoftware for a review of the state of the art.

@@

Bibat can be installed on the Windows, Linux or macOS command line on by running
the command `pip install bibat` and then used by running the command `bibat`.
This command triggers an interactive form which prompts the user to select a
range of customisation options. Bibat then creates a new directory containing
code that implements an example analysis, with customisations reflecting the
user's choices. This analysis works immediately, and can be reproduced with
the single command `make analysis` without the need for any further action by
the user: in this sense bibat comes with batteries included.

@fig-workflow illustrates the components of a bibat-based Bayesian workflow and
shows how it proceeds: the project team edits the code components, then runs
`make analysis` triggering creation of the result components. After inspecting
these they repeat the process, leading to a cycle that ultimately results in a
complete, easily reproducible analysis.

@@

As discussed in @gelmanBayesianWorkflow2020, Bayesian workflows are complicated,
featuring plurality, cyclicity and complexity at many levels. As a specialised
Bayesian workflow template, a key goal for bibat was to manage this complexity.
Bibat achieves this aim by separating non-interacting analysis components into
separate, potentially plural modules and by serialising data to files wherever
possible. Prepared datasets, statistical models, inference configurations,
inference results, plots and analyses all have file representations. Fitting
modes, data manipulations and data models are modularised in code through the
use of appropriately structured data classes and functions. Thanks to this
modular approach it is possible to perform small sub analyses individually
and to iteratively expand the analysis by adding components without needing to
consider everything at once. In addition, bibat ensures that there are minimal
restrictions on the components: for example, datasets need not be singular or
tabular and statistical models need not be representable in formula syntax.
Thanks to these accommodations a project team using bibat should typically not
need to foresee the ultimate requirements of their analysis before starting
the project.

Bibat encourages reproducibility mainly by providing a preconfigured makefile
with a target `analysis` that triggers creation of an isolated environment,
installation of dependencies, data preparation, statistical computation and
analysis of results. In this way a bibat analysis can be reproduced on most
platforms using a single command. In particular, this target attempts to install
cmdstan if necessary, using a recipe tailored to the host operating system.
This functionality addresses a common issue where researchers find it difficult
to install Stan, especially on Windows. A second way in which bibat encourages
reproducibility is by providing its Python code in the form of a preconfigured
package. Thanks to the use of modern conventions for specifying dependencies
and configuring tooling, we expect that this will make bibat analyses easier
for other researchers to work with, thereby facilitating replication while also
making initial development more convenient.

Bibat integrates many widely-adopted, open-source and powerful tools to
implement the components of a Bayesian workflow. Bibat projects are written
in modern Python and uses pydantic [@pydanticdevelopersPydantic2022]
and pandera [@niels_bantilan-proc-scipy-2020] for data modelling, Stan
[@carpenterStanProbabilisticProgramming2017] for statistical inference,
cmdstanpy [@standevelopmentteamCmdStanPy2022] for python-Stan interface, arviz
[@kumarArviZUnifiedLibrary2019] for storing and analyzing inferences and sphinx
[@georgbrandlandthesphinxteamSphinx2022] and quarto [@Allaire_Quarto_2022]
for documentation.  Bibat itself is also written in modern Python and uses the
previously mentioned software as well as the popular  popular template tool
copier [@copierdevelopersCopier2024].

To encourage collaborative development of Bayesian workflow projects, Bibat
projects include a preconfigured test environment, continuous integration,
linting and pre-commit hooks. In addition, including documentation as a
first class component of the analysis addresses a common problem in academic
statistics projects where the paper gets out of sync with the code. Bibat is
continuously tested to ensure that it works on the operating systems Linux,
macOS and Windows. Bibat's continuous integration runs a test suite as well as
an end-to-end functional test on all supported Python versions.

@@

Fitting modes in bibat projects take the form of instances of the class
`FittingMode`. Each fitting mode contains a name, a function that fits a
prepared dataset and instructions for how and where to save the results. For
example, the provided prior sampling fitting mode is called "prior", contains
a function that runs MCMC sampling with the `likelihood` input variable set
to `0`, returning a `CmdStanMCMC` instance, and performs MCMC sampling and writes data to the `InferenceData` group
`prior`. This design allows multiple fits to be stored alongside each other
naturally, facilitating many common Bayesian workflow analyses, and can easily
be configured to prevent unnecessary computations.

Bibat provides fitting modes corresponding for prior sampling, posterior
sampling and k-fold posterior sampling. Users can easily add additional fitting
modes by copying these examples, or modify the `FittingMode` class to achieve
even richer functionality.

@@

Other than bibat, there is currently no interactive template that specifically
targets Bayesian workflow projects. There are some templates that arguably
encompass Bayesian workflow as a special case of data analysis project, such
as cookiecutter-data-science [@drivendataCookiecutterdatascience2022], but
these are of limited use compared with a specialised template due to the many
specificities of Bayesian workflow.

@@

Similarly, bibat includes features that not all analyses will need, such as
accommodating plural inferences, fitting modes and datasets. As a result, bibat
is not optimal for projects that are known in advance to be singular. This could
happen, for example, because there is no uncertainty as to the necessary data
manipulations, or because the aim of the project is to maintain and streamline
an existing statistical model rather than to compare it with others.

While inclusion of potentially unneeded features is an important limitation
of bibat, it is worth noting that it is often difficult to predict in advance
what features a Bayesian workflow project will require, and that it can
be time-consuming to re-write a project after belatedly realising that an
unsupported feature is needed. We therefore think there is value in using bibat
as a default.

Another limitation of bibat is that it makes many opinionated choices about
which tools to use. In particular, languages other than Python, inference
frameworks other than Stan and validation frameworks other than Pydantic are
not supported. We think this limitation is best addressed by the development of
additional analysis templates.

@@

In this case bibat was useful because of its ability to implement arbitrary
statistical models, as latent generalised Pareto distributions are not supported
by generic regression packages like brms and bambi. Further, bibat's modular
design made it easier to implement this medium-sized analysis with two datasets,
two models and six inferences.

@@

In @grovesteddySphincter2024, bibat was used to implement a large analysis
of cerebrovascular data from mice, involving two raw datasets, 6 prepared
datasets, 15 models and 15 inferences. Again it was not possible to implement
the analysis using formula-based packages due to the need for custom modelling
approaches to take into account expert knowledge about both the target system
and the measurement apparatus. Due to the number of datasets and models that
needed to be considered, implementing this analysis without using bibat or a
similar structured approach would have required a combination of more time and/
or compromises in software quality.

>>>>

The term "Bayesian workflow" captures the idea that Bayesian statistical
analysis comprises not just inference, but also specific approaches to related
activities like data preparation, model design, diagnosis, debugging and
criticism. This idea can be found in @boxBayesianInferenceStatistical1992
and has recently received increasing scholarly recognition
[@gelmanBayesianWorkflow2020;@grinsztajnBayesianWorkflowDisease2021;
@gabryVisualizationBayesianWorkflow2019]. Software tools now exist that address
most individual aspects of a Bayesian workflow: see @vstrumbelj2024past for a
review of the state of the art.

@@

In order to use bibat, a user must first install the templating library copier
[@copierdevelopersCopier2024], then choose a directory name and run this command
from the command line:

```sh
$ copier copy gh:teddygroves/bibat my-chosen-directory-name
```

This command triggers an interactive form which prompts the user to select
a range of customisation options. The new directory will then be created
if necessary and filled with code that implements an example analysis, with
customisations reflecting the user's choices. This analysis works immediately,
and can be reproduced with the single command `make analysis` without the need
for any further action by the user: in this sense bibat comes with batteries
included.

@fig-workflow illustrates the components of a bibat-based Bayesian workflow and
shows how it proceeds: the project team edits the code components, then runs
`make analysis`, triggering creation of the result components. After inspecting
these they repeat the process, leading to a cycle that ultimately results in a
complete, easily reproducible analysis.

@@

As discussed in @gelmanBayesianWorkflow2020, Bayesian workflows are complicated,
featuring plurality, cyclicity and complexity at many levels. As a specialised
Bayesian workflow template, a key goal for bibat was to manage this complexity.
Bibat achieves this aim by separating non-interacting analysis components into
separate, potentially plural modules and by serialising data to files wherever
possible. Prepared datasets, statistical models, inference configurations,
inference results, plots and analyses all have file representations. Fitting
modes, data manipulations and data models are modularised in code through the
use of appropriately structured data classes and functions. Thanks to this
modular approach it is possible to perform small sub analyses individually
and to iteratively expand the analysis by adding components without needing to
consider everything at once. In addition, bibat ensures that there are minimal
restrictions on the components: for example, datasets need not be singular or
tabular, and it is possible to use any statistical model that Stan can compile.
Thanks to these accommodations a project team using bibat should typically not
need to foresee the ultimate requirements of their analysis before starting
the project.

Bibat encourages reproducibility by providing a preconfigured makefile
with a target `analysis` that triggers creation of an isolated environment,
installation of dependencies, data preparation, statistical computation and
analysis of results. In this way a bibat analysis can be reproduced on most
platforms using a single command. In particular, this target attempts to install
cmdstan if necessary, using a recipe tailored to the host operating system.
This functionality addresses a common issue where researchers find it difficult
to install Stan, especially on Windows. A second way in which bibat encourages
reproducibility is by providing a preconfigured Python project following modern
conventions, making bibat analyses straightforward to replicated and extend for
other researchers who are already familiar with these conventions.

Bibat integrates many widely-adopted open-source tools to implement
the components of a Bayesian workflow. These include pydantic
[@pydanticdevelopersPydantic2022] and pandera [@niels_bantilan-proc-scipy-2020]
for data modelling, Stan [@carpenterStanProbabilisticProgramming2017] for
statistical inference, cmdstanpy [@standevelopmentteamCmdStanPy2022] for
python-Stan interface, arviz [@kumarArviZUnifiedLibrary2019] for storing and
analysing inferences and sphinx [@georgbrandlandthesphinxteamSphinx2022] and
quarto [@Allaire_Quarto_2022] for documentation.

To encourage collaborative development of Bayesian workflow projects, Bibat
projects include a preconfigured test environment, continuous integration,
linting and pre-commit hooks. In addition, bibat includes documentation as a
first class, integrated component of the analysis, helping to keep it in sync
with the other components.

Bibat is continuously tested to ensure that it works on the operating systems
Linux, macOS and Windows. Bibat's continuous integration runs a test suite as
well as an end-to-end functional test on all supported Python versions.

@@

Fitting modes in bibat projects take the form of instances of the class
`FittingMode`. Each fitting mode contains a name, a function that fits a
prepared dataset and instructions for how and where to save the results. For
example, the provided prior sampling fitting mode is called "prior", contains
a function that runs MCMC sampling with the `likelihood` input variable set to
`0`, returning a `CmdStanMCMC` instance, and specifies that this result should
be written to the `InferenceData` group `prior`.  Bibat provides fitting modes
corresponding for prior sampling, posterior sampling and k-fold posterior
sampling. Users can easily implement additional fitting modes or modify the
`FittingMode` class to achieve even richer functionality. Fitting modes can be
referenced by name from the file that configures an inferece: for example, the
following lines indicate that the inference should be run in prior, posterior
and kfold modes:

```toml
modes = ["prior", "posterior", "kfold"]
```

Fitting modes allow bibat projects to succinctly but flexibly declare how to
perform inferences, and allow results corresponding to the same inference to be
stored alongside each other appropriately.

@@

Other than bibat, we are not aware of any interactive template that specifically
targets Bayesian workflow projects. There are some templates that arguably
encompass Bayesian workflow as a special case of data analysis project,
such as cookiecutter-data-science [@drivendataCookiecutterdatascience2022],
but these are of limited use compared with a specialised template due to
the many specificities of Bayesian workflow. cookiecutter-cmdstanpy-wrapper
[@wardWardBrianCookiecuttercmdstanpywrapper2024] is an interactive template that
targets a different use case than Bayesian workflow projects, namely setting up
a Python package that provides pre-compiled Stan models.

@@

Similarly, bibat accommodates plural inferences, fitting modes and datasets,
but many analyses are singular in at least one of these components and could
therefore be implemented more simply and concisely than an equivalent bibat
project. On the other hand, it is typically difficult to predict in advance
which components of a Bayesian workflow will be plural, and costly to re-write
a project after mistakenly assuming that a component will be singular. While we
acknowledge that accommodating potentially unneeded plurality is an important
limitation of bibat, we nonetheless think that it is the correct choice for a
general-purpose template.

Another limitation of bibat is that it makes many opinionated choices about
which tools to use. In particular, languages other than Python, inference
frameworks other than Stan and validation frameworks other than Pydantic
are not supported. We think that it is on the whole good for templates to be
opinionated, as unopinionated templates are necessarily more complicated; this
limitation of bibat is therefore best addressed by the development of additional
analysis templates that make different choices.

@@

In this case bibat was useful because of its ability to implement arbitrary
statistical models, as latent generalised Pareto distributions are not supported
by any available formula-based regression packages. Further, bibat's modular
design made it easier to implement this medium-sized analysis with two datasets,
two models and six inferences.

@@
